---
title: "Bikeshare in LA, NYC and Phili"
subtitle: "Bikesharing"
author: "Rachel, Ryan, Ina and Gorish"
institute: "Marymount"
date: "`r Sys.Date()`"
output:
 xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Introduction
The first thing we have to do is setup our environment and load our data from bikesharing in LA, NYC and Philadelphia

```{r load-pkg, message = FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
library(readr)
library(readxl)
```

```{r load-data la, message=FALSE, warning=FALSE}
trips <- read_csv("metro-trips-2020-q4.csv")
```

```{r load-data nyc, message = FALSE}
NYtripdata19 <- read_csv("JC-201909-citibike-tripdata.csv")
```

```{r load-data, message = FALSE}
NYtripdata20 <- read_csv("JC-202009-citibike-tripdata.csv")
```

```{r Load-data phili, message=FALSE, warning=FALSE}
philly_trips <- read_excel("philly trips.xlsx")
```

---
```{r message=FALSE, warning=FALSE}
parse_date_time("04-16-2021,1:09:20", orders = "mdy hms" )
```

---
# Let's Look at the LA and Phili Data
```{r}
names(trips)
```
---
# Let's Look at the Data (2)
```{r}
glimpse(trips)
```
---
# Let's Look at the  NYC Data
Data for NYC was from CitiBike and has different fields
```{r}
names(NYtripdata19)
```
---
# Let's Look at the Data (2)
```{r}
glimpse(NYtripdata19)
```
---

# Statistics and visualisation for September 2019 based on NY19 dataframe

```{r creating-variable-age-for-September-2019-assigning-it-to-NY19}
NY19 <- NYtripdata19 %>% 
  mutate (age = (2019-`birth year`))
```


```{r creating-variable-agegroup-for-September-2019-assigning-it-to-NY19}
NY19 <- NY19 %>% mutate(agegroup = case_when(age >= 61 & age <= 131 ~ '4',
                                             age >= 46  & age <= 60 ~ '3',
                                             age >= 30  & age <= 45 ~ '2',
                                             age >= 16  & age <= 29 ~ '1'))

```

```{r assigning-the-gender-variable-names-September-2019}
NY19$gender     <- as.factor(NY19$gender)
levels(NY19$gender) <- c("UNKNOWN", "MALE", "FEMALE")
```
---
```{r plotting-the-histogram-of-the-birth-year-for-September-2019}
ggplot(NY19) + geom_histogram(aes(`birth year`))
```

---
# Slide Title

```{r calculate-statistics-for-age_category-September-2019 NY19}
NY19 %>%  
 summarise(minimum_age=min(age),mean_age=mean(age),median_age=median(age), 
            stdev_age=sd(age))
```
---

```{r histogram-for-tripduration-in seconds-September-2019}
NY19$tripduration <- as.integer(NY19$tripduration)
ggplot(NY19) + geom_histogram(aes(tripduration))
```

---
# Slide Title

```{r histogram-adjusted-x-axis-for-tripduration-in seconds-September-2019 }
x.max <- quantile(NY19$tripduration, 0.99)
ggplot(NY19) + geom_histogram(aes(tripduration)) + xlim(c(0, x.max))
```

```{r creating- new-variable-tripduration-in-minutes-September-2019}
NY19  <- mutate(NY19, tripduration.min = tripduration/60)
x.max <- quantile(NY19$tripduration.min, 0.99)
```
---
```{r defining-tripduration.min-as-integer-September-2019}
NY19$tripduration.min <- as.integer(NY19$tripduration.min)
```

```{r histogram-of-the-tripduration-in minutes-September-2019, message=FALSE, warning=FALSE}
ggplot(NY19) + geom_histogram(aes(tripduration.min, binwidth=2), fill="darkgreen", alpha=0.4) + xlim(c(0, x.max)) + theme_classic()
```





